Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-07-26 GWLDCI.tex
things we fixed in today's meetings

In \cite{wilsonloop} Agarwala and Amat showed that every weakly admissible Wilson loop diagrams can be identified with a type of matroid called a positroid, and observed that this mapping is neither one-to-one nor onto. Each such positroid corresponds to a positroid cell in the appropriate positive Grassmannian, $\Gr(k,n)$ \cite{Postnikov}\sanote{Is this the right citation?}. In this paper we give a simple characterization of this lack of injectivity in terms of an equivalence relation on the Wilson loop diagrams, and enumerate the number of different Wilson loop diagrams that correspond to each positroid cell. We also study the lack of surjectivity, and show that the number of distinct positroid cells associated to Wilson loop diagrams of a given size is given by counting the directions in an associahedron.

@@

Meanwhile, there is a separate body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2011pv, Britto:2005fq, Cachazo:2004kj}. These integrals are calculated via Wilson loops. As noted above, Agarwala and Marin-Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that are realized as an element of the positive Grassmannian $\Gr(k,n)$ \cite{wilsonloop}. In (separate) papers with Fryer and Marcott, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmannians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{Amplituhedronsquared, HeslopStewart}.

@@

\sanote{Everything from here applies to weakly admissible. In fact, I think the dfn of admissible in my paper with Amat is what we call weakly admissible here. Should we change to weakly?}

@@

For an admissible Wilson loop diagram $W$, every triangulated piece $t$ of $\tau(W)$ corresponds to a subdiagram of $W$ obtained by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $W_P = (P, V(P))$.  Conversely, given a subdiagram $W_P = (P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, as follows:
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ whose end points are elements of $V(P)$ which are cyclically consecutive in $[n]$. (This is significantly easier to interpret in picture form: see Example~\ref{eg:subtrees} below.)
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$,
  \item the outer edges of $\tau(W)$ for which both their end points are in the vertex set.
  \end{itemize}
\end{itemize}

@@

The key point to note is that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in \hlfix{the subdiagram $(P,V(P))$, which would have an outer edges between vertices 9 and 1, and between 5 and 8}{Sian replaced $W_P$ with $(P,V(P))$ here and now isn't sure if that was right because this sentence doesn't make sense?}.  In particular, $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$ (as it may also include some of the outer edges of $\tau(W)$), nor is it the polygon dissection of $(P,V(P))$ viewed as a Wilson loop diagram in its own right (which would have vertex set equal to $V(P)$).
\end{eg}

@@

We begin by proving some useful facts about propagator flats, and flats of matroids associated to admissible Wilson loop diagrams more generally. Recall from Definition~\ref{VPropdfn} that the propagator flat $F(P)$ of $P \subseteq \cP$ denotes the set $V(P^c)^c$ of vertices that only support propagators in $P$ (or are non-supporting).\note{Sian: I know we discussed the phrasing of this last time so I'm highlighting this change: I replaced ``(or no propagators)'' with ``(or are non-supporting)'' because it was unclear what part of the sentence the ``no propagators'' was attached to. Is this clearer?}

@@

Since \bas \Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c \eas taking complements gives $P \subseteq \Prop(C)^c = Q$. In other words, $(P, V(P))$ is an exact subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is a maximal exact subdiagram, $P=Q$, or $\Prop (C) = P^c$, and thus $S = \emptyset$, forcing $F(P^c)$ to be a cyclic flat. \sanote{the previous proof of (2) (commented out) was making my head spin. I think this is more straightforward, butI might just be rearranging deckchairs.}
\begin{comment}For part (2), suppose that $(P, V(P))$ is a maximal exact subdiagram. By Lemma \ref{lem decompose flat} (1) and Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}V(P)^c = F(P^c) = C \sqcup S = F(\Prop(C)) \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. We show that $\Prop(C) = P^c$, thus forcing $S$ to be empty. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and not independent (by part (1)).

By similar arguments to Lemma~\ref{lem decompose flat}, the fact that $S$ is independent and that $C$ is the union of all circuits in $F(P^c)$ implies that every element of $S$ is independent of $C$, and therefore 
\[{\rk\big(F(P^c)\big) = \rk(C) + \rk(S)}.\]% this equation doesn't really need to be displayed, but it was causing a badbox 
We also know that $\rk(S) = |S|$, by the independence of $S$. Combining these with Lemma~\ref{lem facts about WLD matroids}(1), we get \ba |S| = \rk(S) = \rk\big(F(P^c)\big) -  \rk(C) \leq |P^c| - \rk(C) \;. \label{increasingrank} \ea

Define $Q = \Prop(C)^c$. We will show that $Q=P$, which will imply that $S=\emptyset$ and that ${F(P^c) = F(Q^c) = F(\Prop(C))}$ is a cyclic flat.

From the definition of $Q$ we obtain $C = F(Q^c)$, and, by Lemma \ref{lem facts about WLD matroids}(2), it follows that $\rk(C) = |Q^c|$. By equation \eqref{increasingrank}, we obtain
\ba |P| + |S| \leq |P| + |P^c| - \rk(C) = |\cP| - |Q^c| = |Q| \;. \label{Qsizebound}\ea 
Meanwhile, from equation \eqref{eq decompose F} we have
\bas V(P) \sqcup S = C^c = F(\Prop(C))^c = F(Q^c)^c = V(Q) \;. \eas
Combining this with equation \eqref{Qsizebound} gives 
\ba |V(Q)| = | V(P) | + |S| = | P | + 3 + |S| \leq |Q| +3 \;. \label{Q defines exact}\ea
Since $W$ is admissible this cannot be a strict inequality, so we have $|V(Q)| = |Q| +3$ and the subdiagram $(Q, V(Q))$ is exact.

Finally, note that since $F(P^c) = C \sqcup S$ (equation~\eqref{eq decompose F}), we have
\[\Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c.\]
Taking complements, we obtain $P \subseteq \Prop(C)^c = Q$, and hence that $(P, V(P))$ is a subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is maximally exact by hypothesis, and $(Q, V(Q))$ is exact by equation \eqref{Q defines exact}, we must have $P = Q$. Therefore $S = \emptyset$ and $F(P^c)$ is a cyclic flat.
\end{comment}

@@

In \cite{wilsonloop}, the authors show that each weakly admissible Wilson loop diagrams correspond to a postiroid. That is, each diagram correspond to matroids that can be represented by elements of the positive Grassmannians $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannians $\Gr(k, n)$, namely the points which represent it \cite{Postnikov}. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.  In other words, there is a map from weakly admissible Wilson loop diagrams, $W$, to matroids, $M(W)$, to positroid cells in the CW complex on $\Gr(|\cP|, n)$.
\note{How is this a def as it doesn't define the cells?  Also, should it be cell singular} \sanote{got rid of dfn}

@@

For the converse, \note{I don't want to mess with this because it was one of the delicate proofs, but why can't we just say $M(W)=M(W')$ so the set of uniform submatroids is the same for $M(W)$ and $M(W')$ so by the previous result the set of exact subdiagrams of $W$ and $W'$ is the same, and all this was inclusion preserving so the set of maximal exact subdiagrams is also the same.} suppose that $M(W) = M(W') = M$. This immediately implies that $|\cP| = |\cP'|$, so to demonstrate the equivalence of $W$ and $W'$ it suffices to show that for any maximal exact subdiagram $(P,V(P))$ in $W$, there is a corresponding maximal exact subdiagram $(P',V(P'))$ in $W'$ with $V(P) = V(P')$.

@@

We know from examples (e.g. \cite{casestudy}) that \hlfix{even after taking the dimension of the positroid cell into consideration}{Sian: I'm not quite sure what ``taking the dimension into consideration'' means here: do we mean ``even after restricting our attention to cells of dimension $3k$''?}\sanote{yes, but it may be worth just getting rid of this text.} \footnote{it is proven in \cite{generalcombinatoricsII} and \cite{Marcott} that Wilson loop diagrams with $k$ propagators map to $3k$ dimensional positroid cells}, many positroid cells do not appear as admissible Wilson loop diagrams with a fixed number of propagators and vertices. That is, the map from admissible Wilson loop diagrams to positroid cells of the correct dimension in the correct Grassmannian neither injective nor surjective. We do not obtain a formula for the number of cells, but do obtain an alternate characterization of the equivalence classes which is the subject of the next section.

@@

Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension. \hlfix{Any two vertices are trivially parallel to each other.}{changed. Correct?}

@@

This proof is by a symmetrical construction, where we give a proof with respect to a single edge in the face $f$, leaving the rest of the argument as read due to the symmetry.\note{Actually, since it is by contradiction we don't need to do both directions; one direction already gives a contradiction and that's enough} \sanote{agreed. Withdraw previous questions}

>>>>

In \cite{wilsonloop} Agarwala and Amat showed that every weakly admissible Wilson loop diagrams can be identified with a type of matroid called a positroid, and observed that this mapping is neither one-to-one nor onto. Each such positroid corresponds to a positroid cell in the appropriate positive Grassmannian, $\Gr(k,n)$ \cite{Postnikov}. In this paper we give a simple characterization of this lack of injectivity in terms of an equivalence relation on the Wilson loop diagrams, and enumerate the number of different Wilson loop diagrams that correspond to each positroid cell. We also study the lack of surjectivity, and show that the number of distinct positroid cells associated to Wilson loop diagrams of a given size is given by counting the directions in an associahedron.

@@

Meanwhile, there is a separate body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2011pv, Britto:2005fq, Cachazo:2004kj}. These integrals are calculated via Wilson loops. As noted above, Agarwala and Marin-Amat uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that are realized as an element of the positive Grassmannian $\Gr(k,n)$ \cite{wilsonloop}. In separate papers with Fryer and Marcott, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmannians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{Amplituhedronsquared, HeslopStewart}.

@@

@@

For an admissible Wilson loop diagram $W$, every triangulated piece $t$ of $\tau(W)$ corresponds to a subdiagram of $W$ obtained by taking the set of propagators $P$ corresponding to edges of $t$ and then taking the subdiagram $(P, V(P))$.  Conversely, given a subdiagram $(P, V(P))$ of $W$ we can obtain a subgraph of $\tau(W)$, called $t$, as follows:
\begin{itemize}
\item The vertex set of the subgraph $t$ is
  \begin{itemize}
  \item the vertices of $\tau(W)$ corresponding to edges of $W$ whose end points are elements of $V(P)$ which are cyclically consecutive in $[n]$. (This is significantly easier to interpret in picture form: see Example~\ref{eg:subtrees} below.)
  \end{itemize}
\item The edge set of the subgraph is
  \begin{itemize}
  \item the edges of $\tau(W)$ corresponding to propagators of $P$,
  \item the outer edges of $\tau(W)$ for which both their end points are in the vertex set.
  \end{itemize}
\end{itemize}

@@

The key point to note is that the subgraph $t$ depends on how the propagators of $P$ sit inside $[n]$, not on how they sit in the subdiagram $(P,V(P))$, which would have outer edges between vertices 8 and 9, 9 and 1, 4 and 5, and 5 and 8.  In particular, $t$ is not the subgraph of $\tau(W)$ consisting only of edges corresponding to propagators of $P$ (as it may also include some of the outer edges of $\tau(W)$), nor is it the polygon dissection of $(P,V(P))$ viewed as a Wilson loop diagram in its own right (which would have vertex set equal to $V(P)$).
\end{eg}

@@

We begin by proving some useful facts about propagator flats, and flats of matroids associated to admissible Wilson loop diagrams more generally. Recall from Definition~\ref{VPropdfn} that the propagator flat $F(P)$ of $P \subseteq \cP$ denotes the set $V(P^c)^c$ of vertices that only support propagators in $P$ or are non-supporting.

@@

Since \bas \Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c \eas taking complements gives $P \subseteq \Prop(C)^c = Q$. In other words, $(P, V(P))$ is an exact subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is a maximal exact subdiagram, $P=Q$, or $\Prop (C) = P^c$, and thus $S = \emptyset$, forcing $F(P^c)$ to be a cyclic flat. \sanote{the previous proof of (2) (commented out) was making my head spin. I think this is more straightforward, butI might just be rearranging deckchairs.}
%\begin{comment}For part (2), suppose that $(P, V(P))$ is a maximal exact subdiagram. By Lemma \ref{lem decompose flat} (1) and Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
%\begin{equation}\label{eq decompose F}V(P)^c = F(P^c) = C \sqcup S = F(\Prop(C)) \sqcup S,\end{equation}
%where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. We show that $\Prop(C) = P^c$, thus forcing $S$ to be empty. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and not independent (by part (1)).
%
%By similar arguments to Lemma~\ref{lem decompose flat}, the fact that $S$ is independent and that $C$ is the union of all circuits in $F(P^c)$ implies that every element of $S$ is independent of $C$, and therefore 
%\[{\rk\big(F(P^c)\big) = \rk(C) + \rk(S)}.\]% this equation doesn't really need to be displayed, but it was causing a badbox 
%We also know that $\rk(S) = |S|$, by the independence of $S$. Combining these with Lemma~\ref{lem facts about WLD matroids}(1), we get \ba |S| = \rk(S) = \rk\big(F(P^c)\big) -  \rk(C) \leq |P^c| - \rk(C) \;. \label{increasingrank} \ea
%
%Define $Q = \Prop(C)^c$. We will show that $Q=P$, which will imply that $S=\emptyset$ and that ${F(P^c) = F(Q^c) = F(\Prop(C))}$ is a cyclic flat.
%
%From the definition of $Q$ we obtain $C = F(Q^c)$, and, by Lemma \ref{lem facts about WLD matroids}(2), it follows that $\rk(C) = |Q^c|$. By equation \eqref{increasingrank}, we obtain
%\ba |P| + |S| \leq |P| + |P^c| - \rk(C) = |\cP| - |Q^c| = |Q| \;. \label{Qsizebound}\ea 
%Meanwhile, from equation \eqref{eq decompose F} we have
%\bas V(P) \sqcup S = C^c = F(\Prop(C))^c = F(Q^c)^c = V(Q) \;. \eas
%Combining this with equation \eqref{Qsizebound} gives 
%\ba |V(Q)| = | V(P) | + |S| = | P | + 3 + |S| \leq |Q| +3 \;. \label{Q defines exact}\ea
%Since $W$ is admissible this cannot be a strict inequality, so we have $|V(Q)| = |Q| +3$ and the subdiagram $(Q, V(Q))$ is exact.
%
%Finally, note that since $F(P^c) = C \sqcup S$ (equation~\eqref{eq decompose F}), we have
%\[\Prop(C) \subseteq \Prop(F(P^c)) \subseteq P^c.\]
%Taking complements, we obtain $P \subseteq \Prop(C)^c = Q$, and hence that $(P, V(P))$ is a subdiagram of $(Q, V(Q))$. Since $(P, V(P))$ is maximally exact by hypothesis, and $(Q, V(Q))$ is exact by equation \eqref{Q defines exact}, we must have $P = Q$. Therefore $S = \emptyset$ and $F(P^c)$ is a cyclic flat.
%\end{comment}

@@

In \cite{wilsonloop}, the authors show that each weakly admissible Wilson loop diagrams correspond to a positroid. That is, each diagram corresponds to a matroid that can be represented by elements of the positive Grassmannian $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannian $\Gr(k, n)$, namely the points which represent it \cite{Postnikov}. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.  In other words, there is a map from weakly admissible Wilson loop diagrams, $W$, to matroids, $M(W)$, to positroid cells in the CW complex on $\Gr(|\cP|, n)$.

@@

For the converse, suppose that $M(W) = M(W') = M$. This immediately implies that $|\cP| = |\cP'|$, so to demonstrate the equivalence of $W$ and $W'$ it suffices to show that for any maximal exact subdiagram $(P,V(P))$ in $W$, there is a corresponding maximal exact subdiagram $(P',V(P'))$ in $W'$ with $V(P) = V(P')$.

@@

We know from examples (e.g. \cite{casestudy}) that many positroid cells do not appear as admissible Wilson loop diagrams with a fixed number of propagators and vertices. That is, the map from admissible Wilson loop diagrams to positroid cells of the correct dimension in the correct Grassmannian neither injective nor surjective. We do not obtain a formula for the number of cells, but do obtain an alternate characterization of the equivalence classes which is the subject of the next section.

@@

Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension. Any two vertices are parallel to each other.

@@
